---
title: "Final dta - Data Science"
author: "Andrés Ponce, Ben West, William Lott"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
---

```{r include=FALSE}
#install.packages("WDI")
library(WDI)
#install.packages("dplyr")
library("dplyr")
#install.packages("plyr")
library(plyr)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("plotly")
library(plotly)
#install.packages("DT")
library(DT)
#install.packages("maptools")
library(maptools)
#install.packages("rworldmap")
library(rworldmap)
#install.packages("reshape")
library(reshape)
#install.packages("knitr")
library(knitr)
#install.packages("gganimate")
library(gganimate)
#install.packages("gifski")
library(gifski)
#install.packages("png")
library(png)
#install.packages("magrittr")
library (magrittr)
#install.packages("mice")
library (mice)
#install.packages("VIM")
library(VIM)
#install.packages("factoextra")
library (factoextra)
#install.packages("stats")
library(stats)
#install.packages("crosstalk")
library(crosstalk)
```


```{r}
#Preparting the Data
dat = WDI(indicator=c('NY.GDP.PCAP.KD', "SP.POP.TOTL", "MS.MIL.XPND.GD.ZS"), country='all', extra=TRUE, start=1960, end=2017)
dat <- dat[complete.cases(dat),]
dat2 <- dat
dat2 <- dat2[!duplicated(dat2[,c("country")]),]
country=c('MX','CA','US')

ccodes <- c("V2","1A", "4E", "T4", "XC", "7E", "T7", "EU", "F1", "XE", "XD", "XF", "ZT", "XH", "XI", "XG", "LA", "V3", "XJ", "T2", "XL", "XO", "XM", "XN", "XQ", "T3", "XP", "OE", "S4", "S2", "V4", "V1", "S1", "T5", "ZF", "XT", "1W")
```

```{r}
#create new dataframe
milexp <- dat %>% filter(!iso2c %in% ccodes)

#rename variables
milexp <- rename(milexp, c(NY.GDP.PCAP.KD="GDP", SP.POP.TOTL="Population",MS.MIL.XPND.GD.ZS="Military.Exp")) 
```



 
# Research Question
This project looks to identify and present trends in global military spending over time. In presenting these trends in a highly visual way, the project attempts to make this information easily understandable and accessible to a wide range of audiences. Overall, the goal is to see where the most money is being spent and to what degree these expenditures are similar or different across geography and time.

# Variables

```{r}
as.datatable(formattable(milexp, 
            align =c("l","c","c","c","c", "r"), 
            list(`Indicator Name` = formatter(
              "span", style = ~ style(color = "grey",font.weight = "bold")) 
)))
```

# Hypothesis
Based on a background knowledge of the subject, the main hypothesis is that military expenditures by superpowers, and specifically the United States, far outstrip military expenditures of all other countries. Indeed, it may well be the case that the United States' military expenditures outpace those of the closest competitors combined. What will be interesting, then, will be to see how these change over time and to what degree we can see peaks or valleys in the expenditures during major events. In other words, a secondary hypothesis is that we should be able to correlate noticeable highs or lows with important world events.

# Method
The methods here are very simple. We do simple exploratory analysis of the variables at the country and regional level in order to identify any outstanding trends or notable deviations from norms. In the first instance, we will present a heat map of the world based on military expenditure over several decades to establish a sense of any patterns and trends that we may encounter with more detailed graphs and charts that will appear later. Then, we break down spending by country and by region and look for trends and patterns over the year. Finally, we will attempt a K-means cluster to see if machine learning techniques identify the same regions in the data that we do when we manually code them by our own definitions (i.e. Middle East and North Africa, North America, etc.)

# Analysis
```{r}
#world map
milexp.recent <- milexp[milexp$year>2016,]
milexp.recent <- subset(milexp, !( iso2c=="B8" | iso2c=='8S' | iso2c=='T6' | iso2c=='XU' | iso2c=='Z4' | iso2c=='Z7' | iso2c=='ZG' | iso2c=='ZJ' | iso2c=='ZQ') )
milexpMap <- joinCountryData2Map(milexp.recent, 
                               nameJoinColumn="country",
                               joinCode="NAME")
mapDevice('x11') #create a world shaped window
#plot the map
mapCountryData( milexpMap, 
                nameColumnToPlot='Military.Exp', 
                catMethod='fixedWidth', 
                numCats=100,
                mapTitle = "2017 Military Expenditures (% of GDP)")
```
We first visualized military expenditures on a global scale. What is most striking is that dominant global players, such as China, Russia, and the US do not spend the most on military expenditures as a percent of their GDP; rather, Angola, Eritrea, and Oman spend the most on military. This significant investment in the military is likely due to the fact that these are conflict-affected or prone states. The map also reveals how a country's level of military expenditures depends on its neighbors' military spending. For example, countries clustered together in the Middle East and in southern Africa appear to have similar levels of military expenditures. Finally, the graph highlights how countries that are more economically prosperous and industrialized, like the US and Russia, tend to spend more on their military. 

```{r}

#Density map for Military Expenditure 1960-2010
p<- milexp %>%  filter(year %in% c(1960, 1970, 1980, 2000, 2010)) %>% ggplot(aes(x= Military.Exp, group=as.factor(year), fill= as.factor(year))) + geom_density(alpha=0.4)+ labs(y= "proportion of countries", x= "Military Expenditure as % of GDP", fill= "Year")+ggtitle(paste("Military Expenditure comparison among countries"))

p <- ggplotly(p)
p
```

```{r Code Chunk, fig.width = 8, fig.height = 8}
#Plot map for  Military Expenditure - GDP over time 
tx <- highlight_key(milexp, ~region)
ggt<- ggplot(tx, aes(GDP, Military.Exp, color= region, frame=year))+
geom_point(aes(size= Population, ids= country))
+ggtitle("Animation of Military Expenditure and GDP Over Time")

highlight(ggplotly(ggt))
```

```{r}

p1<- plot_ly(milexp,
             x= ~year,
             y= ~Military.Exp,
             size= ~GDP,
             color = ~region,
             frame= ~year,
             text= ~country,
             hoverinfo= "text",
             type = "scatter",
             mode= "markers",
             showlegend = TRUE) %>% 
  layout(title = "Military Expenditure by year",
         yaxis= list(title= "Military Expenditure"))
p1

#Line plot for military expenditure over time for each country.
g <- highlight_key(milexp, ~country)
continent_filter <- filter_select("filter", "Select a country", g, ~country)

p <- plot_ly(g, showlegend = F) %>%
  group_by(country) %>%
  add_lines(x = ~year, y = ~Military.Exp, color = ~country) %>%
  layout(xaxis = list(title = "")) %>%
  highlight(selected = attrs_selected(showlegend = F))

bscols(continent_filter, p, widths = 12)

```


```{r}
# analyzing regions
milexp %>%
  filter(iso2c == c('8S', 'ZG', 'Z4', "ZQ", "Z7", "ZJ", "XU")) %>%
  ggplot() + geom_line(mapping = aes(year, Military.Exp, 
                                     group=country, color=country)) +
  labs(y = 'Military Expenditures (% of GDP)') +
  ggtitle(paste('Regional Military Expenditure Comparison'))
  
```

```{r}
# analyzing US, China, and India
milexp %>%
  filter(iso2c == c('US', 'IN', 'CN')) %>%
  ggplot() + geom_line(mapping = aes(year, Military.Exp, 
                                     group=country, color=country)) +
  labs(y = 'Military Expenditures (% of GDP)') +
  ggtitle(paste('Military Expenditure comparison of US, India and China'))
```
We next compared military expenditures of China, India, and United States. In total, these countries have recently spent the most on military expenditures. We can see that the US spends the most on military expenditures as a percent of its GDP, but its expenditures do not remain constant over time. It was highest in the 1960s, but then decreased preciptiously around 1970. This makes sense given the mobilization of forces for the Vietnam War, and then the subsequent demobilization that occured mid 1970s. We then see a spike in military spending in the 1980s which occured under the Reagan administration and then decreased until it once again increased after 9/11. India's rate of military expenditure has has hovered around the same levels after it became a more industrialized country in the 1970s. In regards to China, we see that its military expenditures as a percent of its GDP have not risen significantly. Although China has assumed a more dominant global power, its military expenditure is still dwarfed by US's expendiure. However, as we will later explore, what is impressive is how quickly China became one of the world's biggest military spenders. Overall, the changes in military expenditures largely reflect economic growth and/or conflict. 

```{r}
#Create a variable that is raw Military Expenditures and one that is the mean expenditure by percentage and by raw
#expenditures
milexp <- milexp %>% mutate(`Raw expenditure`= GDP*Population*(`Military.Exp`/100)) %>% group_by(year) %>% 
  dplyr::mutate(`Mean expenditure (%)`=mean(`Military.Exp`)) %>% 
  dplyr::mutate(`Mean raw expenditure ($)`=mean(`Raw expenditure`)) %>% 
  dplyr::mutate(`Mean raw difference`=`Raw expenditure`-`Mean raw expenditure ($)`) %>% 
  dplyr::mutate('Mean pctg difference'=`Military.Exp`-`Mean expenditure (%)`) %>% 
  ungroup()
```

```{r}
milexp$`Raw expenditure lag` <- ifelse(milexp$country==lag(milexp$country)&milexp$year==lag(milexp$year)+1,
                     milexp$`Raw expenditure`-lag(milexp$`Raw expenditure`),
                     0)
```

```{r}
#Looking at dot graphs showing changes in top 20 spending over time
ccodes <- c("B8","T6","V2","1A", "Z4", "4E", "T4", "XC", "Z7", "7E", 
            "T7", "EU", "F1", "XE", "XD", "XF", "ZT", "XH", "XI", 
            "XG", "LA", "V3", "ZJ", "XJ", "T2", "XL", "XO", "XM", 
            "XN", "ZQ", "XQ", "T3", "XP", "XU", "OE", "S4", "S2", 
            "V4", "V1", "S1", "8S", "T5", "ZG", "ZF", "XT", "1W")

ordered <- milexp %>% filter(!iso2c %in% ccodes) %>% mutate(`Raw expenditure`=`Raw expenditure`/1000000) %>% 
  group_by(year) %>% top_n(20, `Raw expenditure`) %>% ungroup()
ordered$Status <- case_when(
  ordered$`Raw expenditure lag`>0 ~"Increased",
  ordered$`Raw expenditure lag`<0~"Decreased",
  ordered$`Raw expenditure lag`==0 ~ "No change",
  ordered$`Raw expenditure lag`== NA ~"No change"
)
ordered <- ordered[complete.cases(ordered),]
```


```{r}
#Look at changes over the years in the top countries form 1960 till 2017
cols <- c('Increased'="aquamarine", "Decreased"="light coral", "No change"="gray91")
movingdots <- ggplot(ordered, aes(country, `Raw expenditure`, color = factor(Status))) +
  geom_point(alpha = 0.7, size = 10) +
  scale_color_manual(values = cols) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  #animation setting
  labs(
    title = "Year: {frame_time}",
    x = "Country", 
    y = "Military Expenditure in USD (billions)"
  ) +
  transition_time(as.integer(year)) + #show the Year as integer
  shadow_mark(alpha = 0.9, size = 5) +
  ease_aes("linear")
animate(movingdots, width= 1800, height=600, renderer = gifski_renderer("gganim1.gif"))
```
There are several interesting points to note here. First and foremost is the cear predominance of the United States in military spending over the years. There are no other contenders that even come close to the United States'. The caveat to this, of course, is that the graph does not include the USSR, only the Russian Federation. Therefore, we get a snapshot of Russia's post-Soviet collapse military spending which, while relatively unimposing at the start, increases steadily over time.
The second is the spectacular rise of China. While it doesn't even appear on the chart until the early 1990s, it rapidy rises through the 2010s to become the second-highest military spending in over a decade. 
That China so quickly achieves this is even more impressive given a third notable trend: overall, the military expenditures of different countries are relatively stable over time. Many of the perennial features on the list hardly move over the years and consist most often of grey dots indicating that they have not changed from one year to the next. Some countries, like Japan, feature a slow upwards creep.
The fourth notable trend, then, is that, for the majority of countries, the predominant trend when there is movement seems to be upward. The military-industrial complex is strong and, for the most part, countries that are on the list stay there and only ever change military expenditure to increase it, in the grand scheme of things.
This brings us to the final trend: large changes seem to be correlated with periods of conflict and strife, unsurprisingly. For example, US military spending increases rapidly and enormously in the years following the 9/11 attacks. Kuwait's only appearance on the map is during what appears to be the beginning of the Gulf War in the 1990s. Poland and Oman both appear on the list in more recent times. This can seemingly be explained by increasing regional tensions with Russia and Iran/Saudi Arabia (by way, possibly of Yemen), respectively, that might make them need to increase their military spending to secure themselves. Saudi Arabia, for its part, is relatively stable over time but begins increasingly rapidly in the 2010s, coinciding with the current conflict in Yemen.



```{r}
#Visualize the same information as bar chart
ordered_tidy <- ordered %>% group_by(year) %>% 
  mutate(rank = dense_rank(desc(`Raw expenditure`))) %>% ungroup()
  
  ordered_formatted <- ordered_tidy %>%
  group_by(year) %>%
  # The * 1 makes it possible to have non-integer ranks while sliding
  mutate(rank = rank(-`Raw expenditure`),
         Value_rel = `Raw expenditure`/`Raw expenditure`[rank==1],
         Value_lbl = paste0(" ",round(`Raw expenditure`/1e9))) %>%
  group_by(country) %>% 
  ungroup()
  
  #then do the graph
staticplot = ggplot(ordered_formatted, aes(rank, group = country, 
                                       fill = as.factor(country), color = as.factor(country))) +
  geom_tile(aes(y = `Raw expenditure`/2,
                height = `Raw expenditure`,
                width = 0.9), alpha = 0.8, color = NA) +
  geom_text(aes(y = 0, label = paste(country, " ")), vjust = 0.2, hjust = 1) +
  geom_text(aes(y=`Raw expenditure`,label = Value_lbl, hjust=0)) +
  coord_flip(clip = "off", expand = FALSE) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_reverse() +
  guides(color = FALSE, fill = FALSE) +
  theme(axis.line=element_blank(),
        axis.text.x=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        axis.title.x=element_blank(),
        axis.title.y=element_blank(),
        legend.position="none",
        panel.background=element_blank(),
        panel.border=element_blank(),
        panel.grid.major=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major.x = element_line( size=.1, color="grey" ),
        panel.grid.minor.x = element_line( size=.1, color="grey" ),
        plot.title=element_text(size=25, hjust=0.5, face="bold", colour="grey", vjust=-1),
        plot.subtitle=element_text(size=18, hjust=0.5, face="italic", color="grey"),
        plot.caption =element_text(size=8, hjust=0.5, face="italic", color="grey"),
        plot.background=element_blank(),
        plot.margin = margin(2,2, 2, 4, "cm"))

#animate this
anim = staticplot + transition_states(year, transition_length = 4, state_length = 1) +
  view_follow(fixed_x = TRUE)  +
  labs(title = 'Top Military Expenditure per year : {closest_state}',  
       subtitle  =  "Top 20 Countries",
       caption  = "Military Expenditure per year | Data Source: World Bank Data")

animate(anim, 200, fps = 5,  width = 1200, height = 1000, 
        renderer = gifski_renderer("gganim2.gif"))
```
This 
# Clustering countries 
```{r}
#Creating aggregated data
new <- milexp %>% select(Military.Exp,Population,GDP,country) 

agg = aggregate(new,
                by = list(new$country),
                FUN = mean)
agg$country=NULL

#taking out regions 
agg<- agg[ -c(28,30,45,54,103,117,144), ]

#creating rownames with country names
agg <- agg %>% as.data.frame()
rownames(agg) <- agg$Group.1
agg <- agg[,-1]

```

### Clustering
```{r}
fviz_nbclust(agg, kmeans, method = "wss")
```

```{r}
# I need to agregate values for better cluster interpretation
agg=scale(agg, center = TRUE, scale = TRUE)
k4 <- kmeans(agg, centers = 3, nstart = 25)

#Plotting K-Means plot
fviz_cluster(k4, stand = T, ellipse.type= "convex", data = agg)
```
